---
title: "Homework 4"
author: "Langdon Holmes"
date-modified: "today"
subtitle: "ANOVA"
format: pdf
code-fold: false
code-overflow: wrap
editor: 
  markdown: 
    wrap: 80
---

# 1 Effect Size

## 1.1 Sample Size and Number of Groups from Summary Statistics

$df_{between} = k - 1$, so $2 = k - 1$ and $3 = k$. There are 3 groups.

$df_{within} = N - k$, so $21 = N - 3$ giving $24 = N$. The overall sample size
is 24.

Since the design is balanced, and there are 3 groups, the per-group sample size
is $24/3 = 8$.

## 1.2 Effect Sizes

```{r}
eta.squared <- function(ss_between, ss_within){
  return(ss_between / (ss_between + ss_within))
}

omega.hat <- function(ss_between, ss_within, ms_within, k){
  numerator <- ss_between - (k - 1)*ms_within
  denominator <- (ss_between + ss_within) + ms_within
  return(numerator/denominator)
}

r.squared.adjusted <- function(N, k, eta_squared){
  return(1 - ((N-1)/(N-k))*(1-eta_squared))
}

eta_squared <- eta.squared(660.3, 1107)
cat("Eta Squared:", eta_squared, end="\n")
cat("Omega Hat Squared:", omega.hat(660.3, 1107, 52.7, 3), end="\n")
cat("Adjusted R Squared:", r.squared.adjusted(24, 3, eta_squared))
```

# 2 Power Analysis

```{r}
#| output: false
library("WebPower")
```


## 2.1 Power to detect Omega Squared .15

WebPower needs effect size in terms of $f$, which is related to $\omega^{2}$ by
$f = \sqrt{\frac{\omega^{2}}{1 - \omega^{2}}}$

```{r}
f <- sqrt(0.15 / (1 - 0.15))
# n is the overall sample size, so 10*4
wp.anova(k=4, n=40, f=f)
```

## 2.2 Minimum per-group sample size

```{r}
wp.anova(k=4, f=.4, power=.8)
```

The minimum per-group sample size to detect an effect size of $f=.4$ is
$\lceil \frac{72.17047}{4} \rceil = 19$.

If we are comfortable rounding down a smidge, a per-group sample size of 18
would be fine as well.

# 3 Contrast

## 3.1 Given sample size

$f$ is related to $d$ by $f = \frac{d}{\sqrt{k\sum_{i}c_{i}^{2}}}$, so
$f = \frac{.5}{\sqrt{5(1/4 + 1/4 + 1/9 + 1/9 + 1/9)}}$, which reduces to
$f = \frac{.5}{\sqrt{5(1/2 + 3/9)}}$ and $f = \frac{.5}{\sqrt{5(15/18)}}$, so
$f = .244949$

```{r}
wp.anova(k=5, n=25, f=0.244949, type="two.sided")
```
The power is .2146, which is quite low.

## 3.2 Given power

```{r}
wp.anova(k=5, f=0.244949, power=0.8, type="two.sided")
```

The minimum per-group sample size to detect an effect size of $f=.25$ is
$\lceil \frac{132.9}{5} \rceil = 27$.