---
title: "Homework 1"
author: "Langdon Holmes"
date-modified: "today"
subtitle: "Factor Analysis"
format: pdf
code-fold: false
code-overflow: wrap
editor: 
  markdown: 
    wrap: 80
---

# Question 1

Given that
$\boldsymbol{R} - \boldsymbol{D}_{\psi}=\boldsymbol{UD}_{\ell}\boldsymbol{U}'$ and that $\Lambda$ is one factor solution, provide a formal proof that $\boldsymbol{\Lambda}'\boldsymbol{\Lambda} = \boldsymbol{D}_{\ell}$ using matrix algebra and the information provided in lecture slides:

```{=latex}
\begin{align*}
&= \boldsymbol{\Lambda}'\boldsymbol{\Lambda} \\
&= (\boldsymbol{U}\boldsymbol{D}_{\ell}^{1/2})'\boldsymbol{U}\boldsymbol{D}_{\ell}^{1/2}  & \text{Definition of a factor loading matrix} \\
&= (\boldsymbol{D}_{\ell}^{1/2})'\boldsymbol{U}'\boldsymbol{U}\boldsymbol{D}_{\ell}^{1/2} & \text{Transpose of a product} \\
&= \boldsymbol{D}_{\ell}^{1/2}\boldsymbol{U}'\boldsymbol{U}\boldsymbol{D}_{\ell}^{1/2} & \text{$\boldsymbol{D}_{\ell}$ is symmetric, so equal to its own transpose} \\
&= \boldsymbol{D}_{\ell}^{1/2}\boldsymbol{I}\boldsymbol{D}_{\ell}^{1/2} & \boldsymbol{U}'\boldsymbol{U} = \boldsymbol{I} \text{ because $\boldsymbol{U}$ is orthogonal}\\
&= \boldsymbol{D}_{\ell}^{1/2}\boldsymbol{D}_{\ell}^{1/2} & \text{Multiplication by Identity matrix preserves value} \\
&= \boldsymbol{D}_{\ell} & \blacksquare
\end{align*}
```

# Question 2

Since $\boldsymbol{P}$ is a population correlation matrix, we can follow Case 2 in the slides. We only need to estimate the squared multiple correlations.

```{r}
P <- matrix(c(1.00, .43, .54, .15, .24, .27,
              .43, 1.00, .63, .15, .17, .21,
              .54, .63, 1.00, .18, .24, .33,
              .15, .15, .18, 1.00, .42, .48,
              .24, .17, .24, .42, 1.00, .61,
              .27, .21, .33, .48, .61, 1.00), nrow = 6)
```

$\boldsymbol{P}^{-1}$

```{r}
solve(P) # inverse
```

$\boldsymbol{P} - \boldsymbol{D}_{\psi}$

```{r}
P.1 <- P # copy matrix
SMCs <- (1 - 1 / diag(solve(P))) # squared multiple correlations (SMCs)
diag(P.1) <- SMCs
P.1
```

$\boldsymbol{D}_{\ell}$ for 2 factors.

```{r}
P.1.eigen <- eigen(P.1)

D_ell <-  diag( P.1.eigen$values[1:2] )
D_ell
```

$\boldsymbol{U}$ for 2 factors.

```{r}
U <- P.1.eigen$vectors[,1:2]
U
```

$\hat{\boldsymbol{\Lambda}} = \boldsymbol{U}\boldsymbol{D}_{\ell}^{1/2}$

```{r}
lambda <- U %*% sqrt(D_ell)
lambda
```



## How good were the prior communality estimates?

The prior communality estimates were pretty close. All of them were smaller than the communalities given by $\hat{\boldsymbol{\Lambda}}$, as promised.

```{r}
communalities <- rowSums(lambda^2)
communalities
```

```{r}
SMCs^2 - communalities^2
```




## Confirm that $\boldsymbol{\Lambda}'\boldsymbol{\Lambda} = \boldsymbol{D}_{\ell}$

```{r}
# we round to 8 decimal places so both matrices have the same precision
# this also takes care of very small floating point errors
lambda_prime_lambda = round(t(lambda) %*% lambda, 8)
if(identical(round(D_ell, 8), lambda_prime_lambda)){
  cat('The matrices are equal!')
}
```

## Sanity Check

The first factor loadings (PA1) are all positive here, but they were negative in my manual analysis. Not sure why this would be, but everything else looks good!

```{r}
library(psych)

P.fa <- fa(
  P, # population correlation matrix
  nfactors=2,
  rotate="none", # maybe we will use rotations later...
  fm="pa", # principal factor method
  max.iter=0 # No iteration (because we have the population correlation matrix).
  )

P.fa$communality # communalities
P.fa$loadings # factor loadings: PA1 and PA2
P.fa$values # Eigenvalues of (P - D_psi)
```

