---
title: "Homework 3"
author: "Langdon Holmes"
date-modified: "today"
subtitle: "Non-Parametric Statistics"
format: pdf
code-fold: false
code-overflow: wrap
editor: 
  markdown: 
    wrap: 80
---


```{r}
A <- c(4.09, 4.96, 4.72, 4.64, 3.14, 3.23)
P <- c(3.51, 5.08, 5.48, 4.41, 3.31, 3.44)
C <- c(4.37, 4.98, 4.66, 3.89, 1.61, 3.48)

df <- data.frame(
  group=factor(c(
    rep("A", times=length(A)),
    rep("P", times=length(P)),
    rep("C", times=length(C))
  )),
  value=c(A, P, C))

permute.cor <- function(x, y, n_permutations, method="pearson", lower=FALSE){
  
  distribution <- c()
  
  observed <- cor(x, y, method=method)
  
  for(i in 1:n_permutations){
    y_permuted <- sample(y)
    distribution[i] <- cor(x, y_permuted, method=method)
  }
  
  result = quantile(distribution, c(.05,.95))
  hist(distribution)
  
  if(lower==TRUE){
    p_value = sum(distribution < observed)/n_permutations
  } else {
    p_value = sum(distribution > observed)/n_permutations
  }
  
  return(p_value)
}

largesample.approx <- function(x, y, method="pearson", lower=FALSE) {
  n = length(x)
  r = cor(x, y, method=method)
  if(method=="kendall"){
    r_var = sqrt((4*n + 10)/(9*(n**2 - n)))
  } else{
    r_var = (1/sqrt(n - 1))
  }
  p = pnorm(r, mean = 0, sd = r_var, lower.tail = lower)
  return(p)
}

```

## 1. Compute the Pearson's correlation (r) of P and C. Test this correlation for significance using the standard parametric method, a random permutation method, and a large-sample approximation. To what extent do these tests agree? Interpret the correlation. [9]

```{r}
cor.test(P, C)

cat("Permutation:                p =",
    permute.cor(P, C, 10000, method="pearson"),
    "\nLarge sample approximation: p =",
    largesample.approx(P, C))

```
A Pearson correlation between Patriotism and Conservatism results in $r = 0.703, p = 0.119$. Using this test, we would fail to reject the null hypothesis and conclude that the two traits are not correlated. A permutation-based analysis results in $p \approx .03$, which would provide sufficient evidence to reject the null hypothesis and conclude that the traits are correlated. The large sample approximation results in $p = 0.058$, which is a higher p-value than the permutation analysis but lower than the standard test's p-value. The large sample approximation is also not significant.


## 2. Compute Spearman's rho (rs) for P and C. Test this correlation for significance using a random permutation method and a large-sample approximation. To what extent do these tests agree? Interpret the correlation. [9]

```{r}
cor.test(P, C, method="spearman")

cat("Permutation:                p =",
    permute.cor(P, C, 10000, method="spearman"),
    "\nLarge sample approximation: p =",
    largesample.approx(P, C, method="spearman"))

```
A Spearman correlation between Patriotism and Conservatism results in $r_{s} = 0.886, p = 0.033$. Using this test, we would reject the null hypothesis and conclude that the two traits are correlated. A permutation-based analysis results in $p \approx .008$, which would indicate ample evidence to reject the null hypothesis and conclude the same. The large sample approximation results in $p = 0.024$, which is a higher p-value than the permutation analysis but lower than the standard test's p-value (and also significant).

## 3. Compute Kendall's tau (rτ) for P and C. Test this correlation for significance using a random permutation method and a large-sample approximation. To what extent do these tests agree? Interpret the correlation. [9]

```{r}
cor.test(P, C, method="kendall")

cat("Permutation:                p =",
    permute.cor(P, C, 10000, method="kendall"),
    "\nLarge sample approximation: p =",
    largesample.approx(P, C, method="kendall"))
```
A Kendall correlation between Patriotism and Conservatism results in $\tau = 0.733, p = 0.056$. Using this test, we would fail to reject the null hypothesis and conclude that the two traits are not correlated. However, a permutation-based analysis results in $p \approx .008$ (almost the same as the permuted Spearman p-value), which would indicate ample evidence to reject the null hypothesis and conclude that there is a correlation between Patriotism and Conservatism. The large sample approximation results in $p = 0.019$, which is a higher p-value than the permutation analysis, but still significant.

## 4. Using Spearman's rho as the measure of correlation (r in the formula), determine the extent to which the relationship between patriotism and conservatism can be explained by authoritarianism. [13]

```{r}
library(ppcor)
pcor.test(P, C, A, method="spearman")

# or, the hard way:
cor.partial <- function(x, y, z, method = "spearman"){
  r_xy <- cor(x, y, method = method)
  r_xz <- cor(x, z, method = method)
  r_yz <- cor(y, z, method = method)
  
  numerator <- r_xy - r_xz*r_yz
  
  # add smallest possible value to avoid zero division when permuting later.
  denominator <- sqrt((1-r_xz**2)*(1-r_yz**2)) + .Machine$double.xmin
  
  r_xy_z <-  numerator / denominator 
  return(r_xy_z)
}

cat("Partial Correlation rho =", cor.partial(P, C, A))

```

The partial correlation between Patriotism and Conservatism is not explained by Authoritarianism, $r_{s} = -0.0294, p = 0.963$. 

## Extra credit: Devise a way to test the partial correlation in (4) for significance using permutation in Resampling Stats, use this method, and interpret the results. [4]

```{r}
permute.cor.partial <- function(x, y, ctrl, n_permutations, method="spearman"){
  
  distribution <- c()
  
  observed <- cor.partial(x, y, ctrl, method=method)
  
  for(i in 1:n_permutations){
    y_permuted <- sample(y)
    distribution[i] <- cor.partial(x, y_permuted, ctrl, method=method)
  }
  
  hist(distribution)
  
  p_value = sum(distribution < observed)/n_permutations
  
  return(p_value)
}

cat("Permutation:p =", permute.cor.partial(P, C, A, 10000))
```

Using a resampling strategy, the probability of finding a more extreme rho value is $p \approx 0.5$ which means there is insufficient evidence to reject the null hypothesis. This means that Authoritarianism does not explain the relationship between Patriotism and Conservatism. 

The distribution of rho values with permutations of Conservatism does not look normally distributed, so I may have made a mistake, or there may be a problem with this approach.