---
title: "Homework 2"
author: "Langdon Holmes"
date-modified: "today"
subtitle: "Multilevel Modeling"
format: pdf
code-fold: false
code-overflow: wrap
editor: 
  markdown: 
    wrap: 80
---

# Question 1

Use the "build-up" stepwise strategy of model building to construct a model for the
educational data, using language test scores (LANGPOST) as the dependent variable and
(potentially, depending on how it goes) percentage minority (PERCMINO) and SES
(SES) as predictors. To anchor your analysis, use the null model (random effects
ANOVA) as your simplest model.

```{r}
library(haven)
df <- read_sav("sb.sav")
```


# Question 2

Now use the "tear down" stepwise strategy. To anchor your analysis, use as the most
complex model one with random intercepts, in which SES serves as a level-1 predictor
with random slopes and PERCMINO as a level-2 predictor of both intercepts and slopes.
Do you settle on the same model as in #1? (I'm not leading you; I really don't know!) [15] 


```{r, message = FALSE, warning = FALSE}

```


# Question 3

Now fit the model with a random intercept and a random slope. Use the website to plot
and interpret the significant cross-level interaction effect. Leave the "df" boxes blank, and
remember that the web page does not understand scientific notation (i.e., if you see
0.193383E-02, enter 0.00193383 instead). If Rweb is not working, you can simply copy
and paste the generated code directly into R. Include and interpret: [20] 

## a

Text output (interpret only the "simple intercepts and simple slopes" and "regions
of significance" sections).



## b

A plot of the simple regression of LANGPOST on SES at three conditional
values of PERCMINO: the minimum observed (0%), middle (45%), and
maximum observed (90%). 

## c

A plot of the confidence bands around the simple slope of LANGPOST regressed
on SES. The x-axis of this plot should extend from the minimum to maximum
observed values of PERCMINO.

# Question 4

Einstein allegedly claimed that you never really know a subject until you can explain it to
your grandmother. Please pretend I am your grandmother, and that I just asked you what
group mean centering and grand mean centering are. Explain these concepts to the best
of your ability. Assume that your "grandmother" has no quantitative training, speaks
fluent English, and is genuinely curious. [10]
