---
title: "Homework 11"
author: "Langdon Holmes"
date-modified: "today"
subtitle: "ANOVA"
format: pdf
code-fold: false
code-overflow: wrap
editor: 
  markdown: 
    wrap: 80
---

```{r, message=FALSE, output=FALSE}
library(tidyverse)
library(emmeans)
library(car)
behavior <- read.table("BehaviorRating.dat", header = TRUE)
ptsd <- read.table("PTSDmixed.dat", header = TRUE) %>%
  mutate(Group = factor(Group,
                        labels=c(
                          "Stress Inoculation",
                          "Prolonged Exposure",
                          "standard Support Counseling")
                        )
         )
```

## 1.1 Present outputs, corrections, draw conclusions.

```{r, message=FALSE, warning=FALSE}
behav.lm <- lm(cbind(Parent, Teacher, Self) ~ 1,
                data = behavior)
measures <- as.factor(c("Parent", "Teacher", "Self"))
behav.aov <- Anova(behav.lm,
                   idata=data.frame(measures),
                   idesign=~measures,
                   type=3)
behav.summary <- summary(behav.aov)
```


```{r, echo=FALSE}
cat("\nResponse transformation matrix:\n")
print(behav.summary$multivariate.tests$measures$P) # Uncorrected
cat("\nUncorrected Univariate Test:\n")
print(behav.summary$univariate.tests) # Assumes sphericity
cat("\nMauchly's test:\n")
print(behav.summary$sphericity.tests) # Mauchly's (not violated)
cat("\nCorrected tests:\n")
print(behav.summary$pval.adjustments[,]) # Corrected
```

Mauchly's test of sphericity assesses whether the assumption of sphericity is
violated. A significant p-value is evidence that the assumption has been
violated. In this case, the test is not significant, but the results are not
reliable, so we can use the Greenhouse-Geisser correction anyway. The test is
significant, which indicates that behavior ratings differ between at least one pairing of parents,
teachers, and student self-assessments.

## 1.2 Recover the ANOVA table

```{r}
SS_total <- 140.2 + 1568.8 + 155.1 # = 1,864.1
```

I calculated the total sum of squares above. Mean squares are the sum of squares divided by their respective degrees of freedom. The other values were pulled directly from the Anova output in 1.1.

|             | SS     | df  | MS    | F     |
|-------------|--------|-----|-------|-------|
| Rater       | 140.2  | 2   | 70.12 | 17.18 |
| Subject     | 1568.8 | 19  |       |       |
| Interaction | 155.1  | 38  | 4.08  |       |
| Total       | 1864.1 | 59  |       |       |

## 1.3 Present multivariate tests

```{r, echo=FALSE}
print(behav.summary$multivariate.tests$measures)
```

The multivariate tests produce equivalent results. They are significant, indicating that at least two of the
respondent types (parent, teacher, self) produced different means.

## 1.4 Test pairwise comparisons using Bonferroni

```{r}
emms <- emmeans(behav.lm, ~ rep.meas, adjust="none")
# unadjusted
print(pairs(emms, adjust="none"))
```


```{r}
# Bonferroni
print(pairs(emms, adjust="bonferroni"))
```

Pairwise contrasts between respondent types reveal that there is no significant
difference between parents' and teachers' opinions of a student's behavior, but
students rate their own behavior higher than both parents and teachers.
Students' self-ratings of behavior diverge more from their parents than from
their teachers.

## 2.1 Construct an ANOVA table.

```{=tex}
\begin{table}[h!]
\begin{tabular}{lllll}
\hline
                         & SS      & df      & MS      & F     \\ \hline
Subject                  & 60      & 23      &         &       \\
\hspace{1em}G            & 21      & 2       & 10.5    & 5.65  \\
\hspace{1em}Subject in G & 39      & 21      & 1.86    &       \\
R                        & 36      & 3       & 12      & 12.6  \\
RxS                      & 150     & 69      &         &       \\
\hspace{1em}RxG          & 90      & 6       & 15      & 15.75 \\
\hspace{1em}RxS in G     & 60      & 63      & 0.95    &       \\ \hline
Total                    & 246     & 95      &         &       \\ \hline
\end{tabular}
\end{table}
```


```{r, output=FALSE}
r=4 # Treatments
g=3 # Groups
n=8 # Subjects

# Degrees of Freedom
g*n - 1 # df S = 23
g - 1 # df G = 2
g*(n - 1) # df S(G) = 21
r - 1 # df R = 3
(r - 1)*(n*g - 1) # df RxS = 69
(r - 1)*(g - 1) # df RxG = 6
g*(r - 1)*(n - 1) # df RxS(G) = 63
g*n*r - 1 # df Total = 95

# Sum of Squares
60 - 21 # SS S(G) = 39
90 + 60 # SS RxS is sum of SS for RxG and RxS(G) = 150
60 + 36 + 150 # SS_Total is sum of SS for Subject, Treatment, Interaction = 246

# Mean Squares
21 / 2 # MS G = 10.5
39 / 21 # MS S(G) = 1.857143
36 / 3 # MS R = 12
90 / 6 # MS RxG = 15
60 / 63 # MS RxS(G) = 0.952381

# F Statistics
10.5 / 1.857143 # F for G = 3.4668
12 / 0.952381 # F for R = 1.603796
15 / 0.952381 # F for RxG = 2.246408

# Critical Values
qf(p=.05, df1 = 2, df2 = 21, lower.tail = FALSE) # Group
qf(p=.05, df1 = 36, df2 = 63, lower.tail = FALSE) # Treatment
qf(p=.05, df1 = 6, df2 = 63, lower.tail = FALSE) # Treatment x Group
```

The critical value for group is $F_{2,21,0.05} = 3.47$  
The critical value for treatment is $F_{36,63,0.05} = 1.60$  
The critical value for the interaction is $F_{6,63,0.05} = 2.25$  

All test statistics are above their critical values, so we can conclude that Group and Treatment have a significant interaction effect.



## 2.2 How many groups, subjects, repeated measures?

There are 3 groups because the df for G is 2. There are 4 repeated measures
because the df for R is 3. There are 8 subjects. Since the df for S is 23, we know that
$g \times n = 24$. We already calculated that $g = 3$, so $n=24/3=8$.

## 3.1 Interaction Plot

```{r}
ptsd.lm <- lm(cbind(Pre, Post, Followup) 
              ~ Group,
              data = ptsd)

emmip(ptsd.lm, Group ~ rep.meas) +
  guides(color=guide_legend(title="Treatment")) +
  xlab("Time") +
  ylab("Symptom Severity")
```
## 3.2 One-way ANOVA on pre-treatment measure

```{r}
summary(aov(Pre ~ Group, data=ptsd))
```
There is not a significant difference between treatments at the time of first measurement. This is good because none of the treatments had been applied yet, so we would not expect any differences. I suppose we can conclude that our sample was reasonably well randomized.

## 3.3 Anova to test interaction

```{r}
measures <- as.factor(c("Pre", "Post", "Followup"))
ptsd.aov <- Anova(ptsd.lm,
                   idata=data.frame(measures),
                   idesign=~measures,
                   type=3)
ptsd.summary <- summary(ptsd.aov)
```


```{r, echo=FALSE}
cat("\nUncorrected Univariate Test:\n")
print(ptsd.summary$univariate.tests) # Assumes sphericity
cat("\nMauchly's test:\n")
print(ptsd.summary$sphericity.tests) # Mauchly's (not violated)
cat("\nCorrected tests:\n")
print(ptsd.summary$pval.adjustments[,]) # Corrected
```

The univariate test of the interaction between treatment and time of measurement is significant with G-G correction. Time of measurement interacts with treatment.

## 3.4 Multivariate test of interaction

```{r}
print(ptsd.summary$multivariate.tests$`Group:measures`)
```

The four tests are not equivalent because we now have 2+ levels in both factors, but all of the tests are still significant. Pillai's trace is the recommended choice. We can conclude that there is a significant interaction between time of measurement and treatment in this study. The multivariate test is more reliable than the univariate test, especially for repeated measures over time.

