---
title: "Homework 3"
author: "Langdon Holmes"
date-modified: "today"
subtitle: "Multilevel Modeling"
format: pdf
code-fold: false
code-overflow: wrap
editor: 
  markdown: 
    wrap: 80
---

```{r, message = FALSE, warning = FALSE}
library(haven) # read .sav file
library(lme4)
library(stargazer) # LaTeX tables
library(performance) # ICC
df <- read_sav("../data/popularmv.sav")
```

# 1
In a univariate random intercept, random slope model, test the effect of SEX on student-rated
popularity (POPULAR). Now test the effect of SEX on teacher-rated popularity (TEACHPOP).
Interpret each set of results in isolation. [12]

```{r}
popular.res5 <- lme(popst~0+conss+sexs+const+sext,
                    random=list(school=~0+conss+sexs+const+sext),
                    weights=varIdent(form=~1|dv),
                    data=df,
                    method='ML',
                    control=lmeControl(maxIter=300,msMaxIter=300,niterEM=100,msMaxEval=500)) #increase iterations for convergence
summary(popular.res5)
```



# 2
In a joint multivariate model that allows all intercepts and slopes to covary, but does not allow
the level-1 residuals to covary across variables, assess the same effects you did in (1). Report and
interpret the results. What is better about the analysis in (2) vs. (1)? [12]

# 3
Now test the effect of the level-2 predictor teacher experience (TEXP) on both student-rated
and teacher-rated popularity (without SEX in the model). Report and interpret the results. [12]

# 4
You have now learned at least two ways to formally test the hypothesis that the effects in (3)
are equal (the deviance test and a multiparameter test). Use both of these methods to test the
hypothesis of equal slopes. Report and interpret the results. Are the p-values the same? Close?
Report them to as many decimal places as possible. [12]

# 5 Power Analysis
Using MLPowSim, conduct an a priori power analysis for the slope point estimates in the
following multilevel model, with conjectured parameter values as indicated:


Limit your attention to a potential data set with 40 clusters of size 10 each, and maximum
likelihood (ML) estimation. Assume the predictors are each standard normal both at level-1 and
level-2. What is the approximate power for detecting each slope at   .05 ? Speculate about
why these power estimates are so different even though both population values are .2. [12]

# Extra Credit
Compose a poem (any type: haiku, limerick, sonnet, ballad, free verse... whatever
you like) describing how multilevel modeling makes you feel. [+3]