---
title: "Homework 10"
author: "Langdon Holmes"
date-modified: "today"
subtitle: "ANOVA"
format: pdf
code-fold: false
code-overflow: wrap
editor: 
  markdown: 
    wrap: 80
---


```{r, message=FALSE, output=FALSE, echo=FALSE}
library(dplyr)
library("WebPower")
```

### 1.1 Complete the table (with sphericity assumption). Show your work. Obtain the p-value of the F test.

|             | SS   | df | MS | F |
|-------------|------|----|----|---|
| Treatment   | 150  | 3  | 50 | 5 |
| Subject     | 900  | 9  | 100|   |
| Interaction | 270  | 27 | 10 |   |
| Total       | 1320 |    |    |   |

Sample Descriptives:  
$k = df_{\tau} + 1 = 4$  
$n = df_{\pi} + 1 = 10$  

Treatment:  
$MS_{\tau} = \frac{SS_{\tau}}{df_{\tau}} = \frac{150}{3} = 50$  

Subject:  
$MS_{\pi} = \frac{SS_{\pi}}{df_{\pi}} = \frac{900}{9} = 100$  

Interaction:  
$SS_{\pi\tau} = SS_{T} - SS_{\pi} - SS_{\tau} = 1,320 - 900 - 150 = 270$  
$df_{\pi\tau} = (k-1)(n-1) = 27$  
$MS_{\pi\tau} = \frac{SS_{\pi\tau}}{df_{\pi\tau}} = \frac{270}{27} = 10$  

$F = \frac{MS_{\tau}}{MS_{\pi\tau}} = \frac{50}{10} = 5$  

```{r}
# pf is for p-values
# qf is for critical values
cat(pf(q = 5, df1 = 3, df2 = 27, lower.tail = FALSE),
    qf(p=.05, df1 = 3, df2 = 27, lower.tail = FALSE))
```
$F_{3,27,.05} = 2.96, p = 0.007$  

There is a significant effect of treatment. I don't believe we can say in which direction with the given summary statistics.

### 1.2 How many subjects are there? How many treatments are there?

There are 10 subjects (n) and 4 treatments (k). I showed "calculations" in 1.1.

### 1.3 Calculate $\hat\omega^2$.

$\hat\omega^2 = \frac{SS_{\tau} - (k-1)MS_{\pi\tau}}{SS_T+MS_\pi} = \frac{150 - 3(10)}{1,320+100} = \frac{120}{1,420} = .085$  

### 1.4 Suppose the correction factor for sphericity is $\hat\epsilon = 0.6$, what is the F statistic? What is its sampling distribution? Find p-value.

```{r}
# pf is for p-values
# qf is for critical values
cat(pf(q = 5, df1 = .06*3, df2 = .06*27, lower.tail = FALSE),
    qf(p=.05, df1 = .06*3, df2 = .06*27, lower.tail = FALSE))
```
$F_{\epsilon*3,\epsilon*27,.05} = F_{.18,1.62,.05} = 19.03, p = 0.11$  

With the sphericity adjustment, the main effect of treatment is not significant.

## 2. Consider a within-subject design with 30 subjects repeated measured under 3 conditions.

### 2.1 Suppose a computer program calculated two estimates of the correction factor $\epsilon: 0.6$ and 0.75. Which is Huynh-Feld estimate and which is Greenhouse-Geisser estimate? Calculate the lower bound estimate.

Greenhouse-Geisser tends to underestimate $\epsilon$, so that is likely to result in the lower value of 0.6.

The lower bound estimate is $\epsilon = 1/(k-1) = \frac{1}{3} = 0.\overline{33}$. This is even more conservative than Greenhouse-Geisser.

### 2.2 What is the power of this design to detect $\omega^{2} = 0.1$? Assume $\epsilon = 0.6$ and $\rho = 0.5$. Present screenshots of WebPower.

```{r}
# d = 0.5
omega <- 0.1
rho <- 0.5
f <- sqrt( 3 * omega/(1-omega) * 1/(1-rho) )
print(f)
wp.rmanova(ng=1, # num groups
           n=30, # subjects
           nm=3, # number of measurements
           nscor=0.6, # nonsphericity correction coefficient
           f=f,
           type=1 # 1 = "Within Effect"
           )
```

## 3. In the attachment you find the data from the Brain Area experiment described in HW9. The three levels 1, 2 and 3 of Area are N, A and B brain areas. The three levels 1, 2 and 3 of Lag are 50, 100 and 150ms. Do the following using SPSS or R.

```{r}
df <- read.table("Avoidance.dat", header = TRUE) %>% 
  mutate(
    Area = factor(Area, labels = c("N", "A", "B")),
    Lag = factor(Lag, labels = c("50ms", "100ms", "150ms"), ordered = TRUE)
    )
summary(df)
```
### 3a obtain an interaction plot with brain region as separate lines and Lag on the x axis.

```{r, warning=FALSE}
interaction.plot(
  x.factor = df$Lag,
  trace.factor = df$Area,
  response = df$Time,
  fun = mean,
  ylab = "Time",
  xlab = "Time Lag",
  col = c("pink", "blue", "green"),
  lty = 1, #line type
  lwd = 2, #line width
  trace.label = "Area"
  )
```

### 3b obtain ANOVA table.

```{r}
fit <- aov(Time ~ Area*Lag, data=df)
summary(fit)
```

### 3c obtain the unadjusted p-value for the interaction contrast in Problem 3 of HW9

|              | Area N | Area A | Area B | Row Means |
|--------------|--------|--------|--------|-----------|
| 50ms         | 28.6   | 16.8   | 24.4   | 23.27     |
| 100ms        | 28     | 23     | 16     | 22.33     |
| 150ms        | 28     | 26.8   | 26.4   | 27.07     |
| Column Means | 28.2   | 22.2   | 22.27  | 24.22     |

```{r}
library(emmeans)

emms <- emmeans(fit, specs = ~ Area*Lag, adjust="none")
interaction_contrasts <- contrast(emms,
                                  list(Lag=c(1, -1, 0),
                                       Area=c(0, 1, -1)),
                                  interaction="consec")
print(interaction_contrasts[2])
```

### 3d obtain unadjusted two-sided p-values for the family of 6 pairwise comparisons in the simple effects of Area (those in Problem 2 of HW9).

```{r}
pairwise_contrasts <- pairs(emms, simple="Area", adjust="none")
pairwise_contrasts[-c(3,6,9)]
```
